Viscous Transport in Eroding Porous Media

### 2019-10-18 docs/transport/erosion_transport.tex
moved a directory
add modified figs in transport paper
add two bodies in caption of transport paper
fixed typos

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Double-Layer Potential Formulation in $\RR^2$}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Applying the same approach as our previous work~\citep{qua-moo2018}, we
start with the double-layer potential 
\begin{align}
  \DDD[\eeta](\xx) = \int_{\bd\Omega} D(\xx,\yy) \eeta(\yy)\, ds_\yy = 
  \frac{1}{\upi}\int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
  \label{eqn:velocityDLP}
\end{align}
where $D$ is the kernel of the integral operator, $\rr = \xx - \yy$,
$\rho = \|\rr\|$, $\nn$ is the unit outward normal at $\yy$, and $\eeta$
is an unknown density function.  We complete the integral equation
formulation by adding the $M$ Stokeslets,
$S[\llambda_\ell,\cc_\ell](\xx)$, and rotlets,
$R[\xi_\ell,\cc_\ell](\xx)$, where $\cc_\ell$ is a point inside the
$\ell^{th}$ body~\citep{pow-mir1987}.  Here $\llambda_\ell$ and
$\xi_\ell$ are the Stokeslet and rotlet strengths, respectively,
corresponding to the $\ell^{th}$ body.  Then, for any sufficiently
smooth geometry $\Omega$, the solution of the incompressible Stokes
equation with a Dirichlet boundary condition $\ff$ is
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:BIE}
\begin{alignat}{3}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
    \NN_0[\eeta](\xx) \nonumber \\
    &\quad + \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx), 
    \quad &&\qquad\xx \in \bd\Omega, \\
  \llambda_\ell &= \frac{1}{2\upi} \int_{\gamma_\ell} 
    \eeta(\yy)\, ds_\yy, &&\qquad \ell = 1,\ldots,M, \\
  \xi_\ell &= \frac{1}{2\upi} \int_{\gamma_\ell}
    (\yy - \cc_\ell)^\perp \cdot \eeta(\yy)\, ds_\yy, 
    &&\qquad \ell = 1,\ldots,M.
\end{alignat}
\end{subequations}
Here, the null space associated with the flux-free condition of $\ff$ is
addressed with  $\NN_0$ which is the integral operator with kernel
$N_0(\xx,\yy) = \nn(\xx) \otimes \nn(\yy)$, $\xx,\yy \in \Gamma$.  In
this work, $\ff$ is the prescribed velocity, which is equal to $\UU$ on
the outer wall, $\Gamma$, and equal to zero on the eroding bodies,
$\gamma_\ell$, $\ell=1,\ldots,M$.

Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, the
corresponding deformation tensor, pressure, and vorticity at $\xx \in
\Omega$ is written in terms of a layer potential~\citep{qua-moo2018}. To
compute the deformation tensor for $\xx \in \gamma$, we include the jump
term
\begin{align}
  \frac{1}{2} \left(\pderiv{\eeta}{\ss} \cdot \ss \right) \left[
    \begin{array}{cc}
      s_x^2 - s_y^2 & 2s_x s_y \\ 2s_x s_y & s_y^2 - s_x^2
    \end{array}
  \right].
  \label{eqn:deformationJump}
\end{align}
Finally, the deformation tensor, pressure, and vorticity due to the
Stokeslets and rotlets are readily available~\citep{poz1992}. Having
computed the deformation tensor on $\gamma$, the shear stress is
computed using equation~\eqref{eqn:shearStress}.

@@

Finally, the Stokes double-layer potential~\eqref{eqn:velocityDLP} can
be written using a Laplace double-layer
potential~\eqref{eqn:laplaceComplex} and its gradients 
\begin{equation}
  \label{eqn:Stokes2Laplace}
  \begin{aligned}
    \DDD[\eeta](\xx) &= 
      \frac{1}{2\upi} \int_{\bd\Omega} 
        \frac{\nn}{\rho^2} (\rr \cdot \eeta) \, ds_\yy + 
      \frac{1}{2\upi} \nabla \int_{\bd\Omega}
        \frac{\rr \cdot \nn}{\rho^2} (\yy \cdot \eeta) \, ds_\yy \\
      &- \frac{1}{2\upi} x_1 \nabla \int_{\bd\Omega}
        \frac{\rr \cdot \nn}{\rho^2}\eta_1(\yy) \, ds_\yy -
      \frac{1}{2\upi} x_2 \nabla \int_{\bd\Omega}
        \frac{\rr \cdot \nn}{\rho^2}\eta_2(\yy) \, ds_\yy.
  \end{aligned}
\end{equation}
Therefore, the Stokes double-layer potential can be written as a sum of
Cauchy integrals and its first derivative~\citep{bar-wu-vee2015}
\begin{equation}
  \begin{aligned}
    u_1(x) &= \Real (v[\psi_1](x)) + \Real (v'[y\cdot\eta](x)) 
             -x_1\Real (v'[\eta_1](x)) - x_2\Real (v'[\eta_2](x)), \\
    u_2(x) &= \Real (v[\psi_2](x)) - \Imag (v'[y\cdot\eta](x)) 
         +x_1\Imag (v'[\eta_1](x)) + x_2\Imag (v'[\eta_2](x)),
  \end{aligned}
  \label{eqn:cauchyVelocity}
\end{equation}
where $y \cdot \eta = y_1 \eta_1 + i y_2 \eta_2 \in \CC$, 
\begin{align} 
  \psi_1=(\eta_1+i\eta_2)\frac{\Real(n)}{n}, \quad
  \psi_2=(\eta_1+i\eta_2)\frac{\Imag(n)}{n},
\end{align}
and $n \in \CC$ is the complex counterpart of the outward unit normal
$\nn \in \RR^2$.

@@

Similar constructions can be used to derive quadrature rules
for $v'(x)$.  For $x \in \Omega$, we have the identity
\begin{align}
  \int_{\bd\Omega} \frac{v^-(y) - v(x) + (y-x)v'(x)}{(y - x)^2} = 0,
\end{align}
and the integrand is bounded for all $x \in \Omega$.  Therefore, after
applying the trapezoid rule and rearranging for $v'(x)$, we have the
interior Barycentric quadrature rule
\begin{align}
  v'(x) = \left(\sum_{j=1}^{N}
    \frac{v^{-}(y_j) - v(x)}{(y_j-x)^2} w_j \right)
  \Bigg/
  \left(\sum_{j=1}^{N} \frac{1}{y_j-x} w_j\right), 
  \quad x \in \Omega.
%  v'(x) = \frac{\sum\limits_{j=1}^{N}
%    \frac{v^{-}(y_j) - v(x)}{(y_j-x)^2} w_j}
%  {\sum\limits_{j=1}^{N} \frac{1}{y_j-x} w_j}, 
%  \quad x \in \Omega.
  \label{eqn:BaryvprimeInterior}
\end{align}
Using a similar construction, the exterior Barycentric quadrature rule
for the the first derivative is
\begin{align}
  v'(x) = \frac{1}{x-a} \left(\sum_{j=1}^N
    \frac{v^+_j - v(x)}{(y_j - x)^2} w_j \right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{(y_j-a)^{-1}}{y_j - x} w_j\right),
    \quad x \in \Omega^c.
%  v'(x) = \frac{1}{x-a} \frac{\sum\limits_{j=1}^N
%    \frac{v^+_j - v(x)}{(y_j - x)^2} w_j}
%    {\sum\limits_{j=1}^N \frac{(y_j-a)^{-1}}{y_j - x} w_j},
%    \quad x \in \Omega^c.
  \label{eqn:BaryvprimeExterior}
\end{align}
Note that $v(x)$ is required to compute $v'(x)$ for both the interior
and exterior case, and this is available using the Barycentric
quadrature rules~\eqref{eqn:BaryvInterior}
and~\eqref{eqn:BaryvExterior}.

@@

\begin{figure}
\begin{center}
  \includegraphics[height=0.227\textwidth]{./figs/20b_dense1}
  \includegraphics[height=0.227\textwidth]{./figs/20b_dense101}
  \includegraphics[height=0.227\textwidth]{./figs/20b_dense201}
  \includegraphics[height=0.227\textwidth]{./figs/20b_dense301}
\caption{\label{fig:Eroding20vort} 20 bodies eroding in a
Hagen-Poiseuille flow.  The four snapshots are equispaced in time, and
the color is the fluid vorticity. In the fourth frame, bodies 4, 12, and
17 have vanished, and body 19 has almost vanished.}
\end{center}
\end{figure}

@@

Finally, we investigate the effect of erosion on pore sizes.  The
distribution of the pore sizes is directly related to the distribution
of the velocity, and thus effects the tortuosity~\citep{den-ica-hid2018}
and anomalous dispersion~\citep{dea-qua-bir-jua2018}. In addition, the
pore sizes are used in network models~\citep{bry-mel-cad1993,
bry-kin-mel1993}.  As described in section~\ref{sec:throats}, we use a
Delaunay triangulation to define neighboring eroding bodies, and we
compute the pore size by finding the closest distance between all
neighboring bodies.  Instead of computing the Delanuay configuration at
each time step, which would result in new definitions for the pores at
each time step, we only compute a new Delaunay triangulation when a
grain completely erodes.  Once all pore sizes are computed, we analyze
their distribution as a function of the porosity.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Double-Layer Potential Formulation in $\RR^2$}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Applying the same approach as our previous work~\citep{qua-moo2018}, we
start with the double-layer potential 
\begin{align}
  \DDD[\eeta](\xx) = \int_{\bd\Omega} D(\xx,\yy) \eeta(\yy)\, ds_\yy = 
  \frac{1}{\upi}\int_{\bd\Omega} 
    \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
    \eeta(\yy) \, ds_\yy, \quad \xx \in \Omega,
  \label{eqn:velocityDLP}
\end{align}
where $D$ is the kernel of the integral operator, $\rr = \xx - \yy$,
$\rho = \|\rr\|$, $\nn$ is the unit outward normal at $\yy$, and $\eeta$
is an unknown density function.  We complete the integral equation
formulation by adding the $M$ Stokeslets,
$S[\llambda_\ell,\cc_\ell](\xx)$, and $M$ rotlets,
$R[\xi_\ell,\cc_\ell](\xx)$, where $\cc_\ell$ is a point inside the
$\ell^{th}$ body~\citep{pow-mir1987}.  Here $\llambda_\ell$ and
$\xi_\ell$ are the Stokeslet and rotlet strengths, respectively,
corresponding to the $\ell^{th}$ body.  Then, for any sufficiently
smooth geometry $\Omega$, the solution of the incompressible Stokes
equation with a Dirichlet boundary condition $\ff$ is
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + 
    \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx), \quad \xx \in \Omega,
\end{align}
where the density function, Stokeslets, and rotlets satisfy
\begin{subequations}
\label{eqn:BIE}
\begin{alignat}{3}
  \ff(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
    \NN_0[\eeta](\xx) \nonumber \\
    &\quad + \sum_{\ell=1}^M S[\llambda_\ell,\cc_\ell](\xx) + 
    \sum_{\ell=1}^M R[\xi_\ell,\cc_\ell](\xx), 
    \quad &&\qquad\xx \in \bd\Omega, \\
  \llambda_\ell &= \frac{1}{2\upi} \int_{\gamma_\ell} 
    \eeta(\yy)\, ds_\yy, &&\qquad \ell = 1,\ldots,M, \\
  \xi_\ell &= \frac{1}{2\upi} \int_{\gamma_\ell}
    (\yy - \cc_\ell)^\perp \cdot \eeta(\yy)\, ds_\yy, 
    &&\qquad \ell = 1,\ldots,M.
\end{alignat}
\end{subequations}
Here, the null space associated with the flux-free condition of $\ff$ is
addressed with  $\NN_0$ which is the integral operator with kernel
$N_0(\xx,\yy) = \nn(\xx) \otimes \nn(\yy)$, $\xx,\yy \in \Gamma$.  In
this work, $\ff$ is the prescribed velocity, which is equal to $\UU$ on
the outer wall, $\Gamma$, and equal to zero on the eroding bodies,
$\gamma_\ell$, $\ell=1,\ldots,M$.

Once~\eqref{eqn:BIE} is solved for the density function $\eeta$, the
corresponding deformation tensor, pressure, and vorticity at $\xx \in
\Omega$ are written in terms of layer potentials~\citep{qua-moo2018}.
To compute the deformation tensor for $\xx \in \gamma$, we include the
jump term
\begin{align}
  \frac{1}{2} \left(\pderiv{\eeta}{\ss} \cdot \ss \right) \left[
    \begin{array}{cc}
      s_x^2 - s_y^2 & 2s_x s_y \\ 2s_x s_y & s_y^2 - s_x^2
    \end{array}
  \right].
  \label{eqn:deformationJump}
\end{align}
Finally, the deformation tensor, pressure, and vorticity due to the
Stokeslets and rotlets are readily available~\citep{poz1992}. Having
computed the deformation tensor on $\gamma$, the shear stress is
computed using equation~\eqref{eqn:shearStress}.

@@

Finally, the Stokes double-layer potential~\eqref{eqn:velocityDLP} can
be written using a Laplace double-layer
potential~\eqref{eqn:laplaceComplex} and its gradients 
\begin{equation}
  \label{eqn:Stokes2Laplace}
  \begin{aligned}
    \DDD[\eeta](\xx) &= 
      \frac{1}{2\upi} \int_{\bd\Omega} 
        \frac{\nn}{\rho^2} (\rr \cdot \eeta) \, ds_\yy + 
      \frac{1}{2\upi} \nabla \int_{\bd\Omega}
        \frac{\rr \cdot \nn}{\rho^2} (\yy \cdot \eeta) \, ds_\yy \\
      &- \frac{1}{2\upi} x_1 \nabla \int_{\bd\Omega}
        \frac{\rr \cdot \nn}{\rho^2}\eta_1(\yy) \, ds_\yy -
      \frac{1}{2\upi} x_2 \nabla \int_{\bd\Omega}
        \frac{\rr \cdot \nn}{\rho^2}\eta_2(\yy) \, ds_\yy.
  \end{aligned}
\end{equation}
Therefore, the Stokes double-layer potential can be written as a sum of
Cauchy integrals and its first derivative~\citep{bar-wu-vee2015}
\begin{equation}
  \begin{aligned}
    u_1(x) &= \Real (v[\psi_1](x)) + \Real (v'[y\cdot\eta](x)) 
             -x_1\Real (v'[\eta_1](x)) - x_2\Real (v'[\eta_2](x)), \\
    u_2(x) &= \Real (v[\psi_2](x)) - \Imag (v'[y\cdot\eta](x)) 
         +x_1\Imag (v'[\eta_1](x)) + x_2\Imag (v'[\eta_2](x)),
  \end{aligned}
  \label{eqn:cauchyVelocity}
\end{equation}
where $y \cdot \eta = y_1 \eta_1 + y_2 \eta_2$, 
\begin{align} 
  \psi_1=(\eta_1+i\eta_2)\frac{\Real(n)}{n}, \quad
  \psi_2=(\eta_1+i\eta_2)\frac{\Imag(n)}{n},
\end{align}
and $n \in \CC$ is the complex counterpart of the outward unit normal
$\nn \in \RR^2$.

@@

Similar constructions can be used to derive quadrature rules
for $v'(x)$.  For $x \in \Omega$, we have the identity
\begin{align}
  \int_{\bd\Omega} \frac{v^-(y) - v(x) + (y-x)v'(x)}{(y - x)^2} = 0,
\end{align}
and the integrand is bounded for all $x \in \Omega$.  Therefore, after
applying the trapezoid rule and rearranging for $v'(x)$, we have the
interior Barycentric quadrature rule
\begin{align}
  v'(x) = \left(\sum_{j=1}^{N}
    \frac{v^{-}(y_j) - v(x)}{(y_j-x)^2} w_j \right)
  \Bigg/
  \left(\sum_{j=1}^{N} \frac{1}{y_j-x} w_j\right), 
  \quad x \in \Omega.
%  v'(x) = \frac{\sum\limits_{j=1}^{N}
%    \frac{v^{-}(y_j) - v(x)}{(y_j-x)^2} w_j}
%  {\sum\limits_{j=1}^{N} \frac{1}{y_j-x} w_j}, 
%  \quad x \in \Omega.
  \label{eqn:BaryvprimeInterior}
\end{align}
Using a similar construction, the exterior Barycentric quadrature rule
for the first derivative is
\begin{align}
  v'(x) = \frac{1}{x-a} \left(\sum_{j=1}^N
    \frac{v^+_j - v(x)}{(y_j - x)^2} w_j \right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{(y_j-a)^{-1}}{y_j - x} w_j\right),
    \quad x \in \Omega^c.
%  v'(x) = \frac{1}{x-a} \frac{\sum\limits_{j=1}^N
%    \frac{v^+_j - v(x)}{(y_j - x)^2} w_j}
%    {\sum\limits_{j=1}^N \frac{(y_j-a)^{-1}}{y_j - x} w_j},
%    \quad x \in \Omega^c.
  \label{eqn:BaryvprimeExterior}
\end{align}
Note that $v(x)$ is required to compute $v'(x)$ for both the interior
and exterior case, and this is available using the Barycentric
quadrature rules~\eqref{eqn:BaryvInterior}
and~\eqref{eqn:BaryvExterior}.

@@

\begin{figure}
\begin{center}
  \includegraphics[height=0.227\textwidth]{./figs/20b_1}
  \includegraphics[height=0.227\textwidth]{./figs/20b_101}
  \includegraphics[height=0.227\textwidth]{./figs/20b_201}
  \includegraphics[height=0.227\textwidth]{./figs/20b_301}
\caption{\label{fig:Eroding20vort} 20 bodies eroding in a
Hagen-Poiseuille flow.  The four snapshots are equispaced in time, and
the color is the fluid vorticity. In the fourth frame, bodies 4, 12, and
17 have vanished, and bodies 7, 19, 20 have almost vanished.}
\end{center}
\end{figure}

@@

Finally, we investigate the effect of erosion on pore sizes.  The
distribution of the pore sizes is directly related to the distribution
of the velocity, and thus effects the tortuosity~\citep{den-ica-hid2018}
and anomalous dispersion~\citep{dea-qua-bir-jua2018}. In addition, the
pore sizes are used in network models~\citep{bry-mel-cad1993,
bry-kin-mel1993}.  As described in section~\ref{sec:throats}, we use a
Delaunay triangulation to define neighboring eroding bodies, and we
compute the pore size by finding the closest distance between all
neighboring bodies.  Instead of computing the Delaunay configuration at
each time step, which would result in new definitions for the pores at
each time step, we only compute a new Delaunay triangulation when a
grain completely erodes.  Once all pore sizes are computed, we analyze
their distribution as a function of the porosity.
